---
title: "Exemplary Code Chunk"
author: "R.Kam"
format: pdf
editor: visual
---
```{r packages}
#| warning: false 
#| message: false

library(tidyverse)
library(knitr)
library(purrr)
library(rvest)
library(palmerpenguins) # To help with better graph visualization
```

# `leg_summary()` — A Reusable Group-Summary Utility in R

A small code-craft exhibit demonstrating a custom tidyverse function (`leg_summary()`) that produces clean grouped summary statistics — counts, means, medians, ranges, totals — for any numeric variable in any tidy dataset. Built in R + Quarto, output as a PDF.

## What this repo is — and isn't

This repo is a **standalone code exhibit**, not a full data project. Its purpose is narrow: showcase one reusable function and demonstrate that it generalizes across datasets via a built-in R example (`palmerpenguins`). The function itself was originally written for a separate web-scraping / Senate legislation project; this exhibit lifts it out of that context to display the code-craft on its own.

If you came here looking for the Senate scraping pipeline, that work lives in a different repository.

## From ideation to execution

### Where this started

The project began as a class assignment with a deceptively simple prompt: *write a custom function in R*. I could have written something trivial — a one-line wrapper around `mean()`, say — but I wanted to write something I'd actually reuse. At the time I was already deep in a separate project scraping U.S. Senate legislative records, and I noticed I was repeatedly typing the same `group_by() %>% summarize()` chain to produce counts, means, medians, ranges, and totals broken out by party or by chamber. Every analysis. Same five statistics. Different grouping variable.

That repetition was the friction point the function was designed to remove.

### Design decisions, and why

A few choices that shaped the final shape of `leg_summary()`:

- **Three positional arguments — `data`, `grouping`, `var` — in that order.** The data-first signature mirrors tidyverse convention so the function pipes naturally. Grouping comes before the summary variable because that's how a human describes the operation aloud: *"summarize this data by party, looking at bills sponsored."*
- **Tidy evaluation via `{{ }}` (curly-curly).** I wanted callers to pass bare column names (`Party`, `body_mass_g`) the way they'd write them inside `dplyr`, not strings (`"Party"`). Curly-curly is the modern tidyverse approach for this; getting it right took some reading on quasi-quotation but it's now an idiom I reach for whenever I write a function that takes column references.
- **The function returns a `kable()` table, not a tibble.** This is opinionated: the caller can't pipe the output further into more `dplyr` operations. But the function exists to produce *human-readable summaries for documents*, not intermediate data, so trading flexibility for slot-and-go formatting was the right call for the use case.
- **Up-front NA filtering on the summary variable.** Rather than dropping `na.rm = TRUE` into every aggregation call, I filter once at the top. Cleaner code, single defined behavior, no risk of one statistic silently behaving differently from another.

### The three things that were genuinely hard

**1. Auto-generating the `kable()` caption from the grouping variable's name.**

I wanted each output table to label itself — `"Summary by Party"`, `"Summary by species"` — without forcing the caller to pass the label in as a separate string argument. Sounds simple. It is not simple. By the time `grouping` arrives inside the function body, it's already been captured as a quosure for tidy evaluation; you can't just paste it into a string. The fix is `deparse(substitute(grouping))`, which extracts the unquoted name as it appeared at the call site. This was the single most stubborn debugging session of the project — the kind of thing that's two lines once you see it and a wall before you do.

**2. Making the function generalize to a totally unrelated dataset.**

A function that "works" on the data you wrote it for isn't proven to be reusable — it's just specific. I picked `palmerpenguins` deliberately for the demonstration: different domain (biology vs politics), different variable types, no overlap in column names with anything legislative. If `leg_summary(penguins, species, body_mass_g)` produced output of the same shape and quality as it did on Senate data, the abstraction was real. It did. That's why the penguins demo is in this exhibit.

**3. Debugging the original Senate data integration.**

The parent project required scraping data with `rvest`, handling sites with anti-scraping protocols, and reshaping the raw HTML into a clean dataframe before the function could touch it. A lot of the function's design — particularly the NA handling and the reliance on tidy column references — got iterated based on real failures hitting it with messy real-world data. That upstream work isn't in this exhibit (it lives in the parent repo), but its fingerprints are all over the function's defensive choices.

### What I'd do differently next time

- **Optional tibble-return mode.** A `format = c("kable", "tibble")` argument would let the function serve both rendering and pipeline use cases.
- **Survey weights.** A `weight` argument would generalize this to weighted survey data, which would have made my Arab Barometer analysis tidier.
- **Input validation.** Right now if `var` isn't numeric, the function fails on the call to `mean()` rather than telling the caller their input is wrong. A two-line `stopifnot()` at the top would make failures legible.

## The function

```{r function}
leg_summary <- function(data, grouping, var) {
  data |>
    filter(!is.na({{ var }})) |>
    group_by({{ grouping }}) |>
    summarize(
      N = n(),
      Mean = mean({{ var }}),
      Median = median({{ var }}),
      Min = min({{ var }}),
      Max = max({{ var }}),
      Total = sum({{ var }})
    ) |>
    kable(
      digits = 2,
      caption = paste("Summary by", deparse(substitute(grouping)))
    )
}
```

## Demonstration

The exhibit calls `leg_summary()` on the `palmerpenguins::penguins` dataset, grouped by species and summarizing body mass (grams). A boxplot follows the table to make the distribution visible alongside the summary statistics — proving the function works on data that has nothing to do with the original use case.

## Additional proof that the code is replicable in our contexts

Using in-built datasets within R, we are able to replicate the code and ensure that it is able to produce the same result that we observe for the Senate data

```{r penguin replicate}
#| warning: false
#| message: false
# Demonstrate using the penguins data within R
leg_summary(penguins, species, body_mass_g)
```

## Visualizing the penguin data using a graph

```{r penguin-graph}
#| warning: false 
#| message: false
#| fig.width: 8 
#| fig.height: 5
penguins |>
  drop_na(species, body_mass_g) |>
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(color = "black", alpha = 0.6) +
  scale_fill_manual(values = c("Adelie" = "darkorange", 
                               "Chinstrap" = "purple1", 
                               "Gentoo" = "dodgerblue3")) +
  labs(
    title = "Penguin Body Mass Distribution by Species",
    subtitle = "A visual demonstration of the native palmerpenguins dataset",
    x = "Penguin Species",
    y = "Body Mass (grams)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") # Hides the redundant legend
```

# Conclusion

The customized function provides a useful tool for summarizing grouped data and can be replicated for similar situations across domains — the demonstration on `palmerpenguins` shows it generalizes beyond its original legislative use case to any tidy dataset with a categorical grouping and a numeric summary variable.

## Acknowledgments

A Gemini tutoring session assisted with debugging the tidy-evaluation syntax in the function — particularly the `deparse(substitute())` pattern for extracting the grouping variable's name at call time. Session transcript: [linked here](https://gemini.google.com/share/a24a1ecf52cc). All design decisions, the demonstration choice, and the framing of this exhibit are the author's own.

## Author

R. Kam · American University, School of International Service
